Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2023-12-20 2024pipit/section/20-Motivation.tex
ok

Pure streaming languages like Lustre lie in an uneasy space somewhere between pure functions and imperative programs.
Like pure functions, pure streaming languages are equational and support reasoning by unfolding and rewriting.
Like imperative programs, they describe how state evolves over time.
Reasoning about them is somewhat tricky, then, as one must reason about finite prefixes of streams.

As a concrete example to introduce Pipit, we consider a program with a static schedule of \emph{triggers}, or actions to be performed at a particular time.
These triggers are stored in a constant array.
The program maintains an index that refers to the current active trigger; at each instant in time, the program checks if the active trigger has expired, and if so, it increments the index.
This trigger logic is a simplification of the time-triggered CAN bus specification \cite{fuehrer2001time}; we will discuss further in \autoref{s:evaluation}.

We first implement a streaming function \emph{count_when} to maintain the index; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
At each time step, $\mbox{count_when}~\textit{max}~\textit{inc}$ checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the previous counter as-is.

% \newcommand\signal{\textit{signal}}
\begin{tabbing}
  @MM@\= @MM@ \= \kill
  @let@ count_when ($\textit{max}$: $\NN$) ($\textit{inc}$: stream $\BB$): stream $\NN$ = \\
    \> $@rec@~\rawbind{\textit{count}}{}$ \\
    \> \> @let@ $\textit{count'}$ @=@ $(0~@fby@~\textit{count}) + (@if@~\textit{inc}~@then@~1~@else@~0)$ @in@ \\
    \> \> $\xcheckP{\PSUnknown}{(0 \le \textit{count} \le \textit{max})}$; \\
    \> \> @if @ $\textit{count'} \ge \textit{max}$ @then@ $\textit{max}$  @else@ $\textit{count'}$ \\
    % \> $\textit{count}$ \\
\end{tabbing}

The implementation of \emph{count_when} first defines a recursive stream, \emph{count}, which in turn defines a local incremented stream \emph{count'}.
The syntax $0~@fby@~\textit{count}$ can be read as zero \emph{followed by} the previous count.

The syntax $\xcheckP{\PSUnknown}{(0 \le \textit{count} \le \textit{max})}$ asserts that the count is within the range $[0, \textit{max}]$.
The subscript $\PSUnknown$ on the check is the \emph{property status}, which in this case denotes that the assertion has been stated, but it is not yet known whether it holds.
A property status of $\PSValid$, on the other hand, denotes that a property has been proved to hold.
These property statuses are used to defer checking properties until enough is known about the environment, and to avoid rechecking properties that have already been proven.

@@

To actually prove the property above, we translate the program into a transition system and prove the property inductively on the system.
Finally, we \emph{bless} the expression, which marks the properties as valid.
Blessing is an intensional operation: it modifies the internal structure of the expression, but does not affect the operational semantics.

@@

The subscript 1 in the translation to transition system and blessing operations refers to the fact that the stream function has one parameter.
This proof does require some boilerplate; in the future we plan to minimise this with metaprogramming.

Callers of count_when can now use the valid variant without needing to re-prove the count-range property.

\begin{tabbing}
  @MM@\= @MM@ \= \kill
  @let@ trigger_fetch ($\textit{index}$: stream $\NN$): stream trigger = \\
    \> $@contract@_{\PSUnknown}$ \\
    \> \> \{ $0 \le \textit{index} \le \textit{maximum}$ \} \\
    \> \> Array.index_const \textit{triggers_array} \textit{index} \\
    \> \> \{ trigger.~$\true$ \} \\
\end{tabbing}

% \section{Pipit in flight}
% \pagebreak
\section{Programming and verifying in Pipit}
\label{s:tut}

\TODO{KILLMERGE}

A common requirement in controllers is to filter an input signal, perhaps using a \emph{finite impulse response} (FIR) filter, which is equivalent to a weighted moving average.
An FIR filter takes a vector of coefficients and an input signal; at every point in time, it computes the dot product of the coefficients and the most recent values of the signal.
We can implement an FIR filter in Pipit as follows:

\newcommand\signal{\textit{signal}}
\begin{tabbing}
  @MM@\= @MMMMMM@ \= \kill
  @let@ fir ($\textit{coefficients}$: list $\RR$) ($\signal$: stream $\RR$): stream $\RR$ = \\
    \> @match@ $\textit{coefficients}$ @with@ \\
    \> @|@ $[]$ \> $\to 0$ \\
    \> @|@ $c :: \cs $ \> $\to (\signal \cdot c) + (\xfby{0}{(\mbox{fir~} \cs~\signal)})$
\end{tabbing}

% Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
The coefficient vector is represented by a list of reals, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
% \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
The implementation starts by looking at the list of coefficients and returns zero if the list is empty.
If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before summing the two parts.
At the start of execution the delay is initially zero.

Pipit is an \emph{embedded} language, like Bedrock~\cite{chlipala2013bedrock}: in this example, the stream type denotes an actual Pipit expression, while the list type and its associated pattern match are part of the \fstar{} meta-language.
To get the real Pipit program, we need to apply the filter to a concrete list of coefficients:

\newcommand\bibo{\mbox{bibo}}
\newcommand\fir{\mbox{fir}}
\newcommand\ii{\textit{input}}
\newcommand\oo{\textit{output}}
\newcommand\ok{\textit{ok}}
\begin{tabbing}
  @MM@\= @MMMMMM@ \= \kill
  @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
    \> $\mbox{fir}~[0.7; 0.3] \ii$
\end{tabbing}

Normalising this definition evaluates away all of the lists.
The result fits in the core language defined in \autoref{s:core}, for which we can generate real-time imperative code:

\begin{tabbing}
    @MM@\= @MMMMMM@ \= \kill
      @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
    \> $(0.7 \cdot \ii) + (\xfby{0}{((0.3 \cdot \ii) + (\xfby{0}{0}))})$
\end{tabbing}

The properties that we want to state about reactive programs usually involve some temporal aspect.
Rather than defining a separate specification language, we implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
We name the past-globally operator \emph{sofar}, as in \emph{the predicate has been true so far}:

\begin{tabbing}
  @MM@\= @MMMMMM@ \= \kill
  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
    \> $\xrec{p'}{p \wedge (\xfby{@true@}{p'})}$
\end{tabbing}

This definition takes a stream of predicates $p$ and introduces a recursive stream $p'$.
At each step, the recursive stream $p'$ checks that the current predicate is true ($p$), and also checks that \emph{sofar} was previously true ($\xfby{@true@}{p'}$).
If there is no previous value, it defaults to true.

\subsection{Bounded input, bounded output}
\label{ss:tut:bibo}
We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range:
% Properties like this can occur as an internal invariant in a larger system:

\begin{tabbing}
  @MM@\= @LET OUTPUT@\= @EXAMPLE1 INPUT@ \= \kill
  @let@ $\mbox{bibo}_2$ (n: $\RRgez$) ($\ii$: stream $\RR$): stream $\BB$ = \\
  \> @check@ $(\mbox{sofar}(|\ii| \le n) \implies |\mbox{fir}_2~\ii| \le n)$
\end{tabbing}

This property states that if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
To prove that this property holds, we translate to a transition system and show that the stream is always true.
In this case, induction over the transition relation is sufficient to prove the property.
There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:

\begin{tabbing}
  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
  @let@ $\mbox{proof}_2$ (n: $\RRgez$): Lemma (induct ($\bibo_2$ n)) = \\
  \> @assert@ (base\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ()); \\
  \> @assert@ (step\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ())
\end{tabbing}

This definition uses \fstar{}'s \emph{lemma} syntax to state that the BIBO property holds inductively for any $n$.
The two assertions prove the inductive cases separately, using our \emph{simplify} tactic to ensure that the translation to transition system is normalised away, and any translation artefacts are removed.

If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is not sufficient: the relationship between the stacked delays in the filter and sofar is not clear from a single step of the transition system.
One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:

\begin{tabbing}
  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
  @let@ $\mbox{proof}_3$ (n: $\RRgez$): Lemma (induct\_k 2 ($\bibo_3$ n)) = \\
  \> @assert@ (base\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ()); \\
  \> @assert@ (step\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ())
\end{tabbing}

Although the properties here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation to verify reactive systems.

>>>>

Pure streaming languages like Lustre lie in an uneasy space somewhere between pure functions and imperative programs.
Like pure functions, pure streaming languages are equational and support reasoning by unfolding and rewriting.
Like imperative programs, they describe how state evolves over time.
Additionally, Lustre programs tend to be composed of many nested recursive loops.
When verifying both pure and imperative programs, recursion or loops generally require explicit loop invariants.
Requiring an explicit invariant for each recursive stream quickly becomes unwieldy for Lustre programs.

As a concrete example to introduce Pipit, we consider a program with a static schedule of \emph{triggers}, or actions to be performed at a particular time.
These triggers are stored in a constant array.
The program maintains an index that refers to the current active trigger; at each instant in time, the program checks if the active trigger has expired, and if so, it increments the index.
This trigger logic is a simplification of the time-triggered CAN bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.

We first implement a streaming function \emph{count_when} to maintain the index; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
At each time step, count_when checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the previous counter as-is.

\begin{tabbing}
  @MM@\= @MM@ \= \kill
  @let@ count_when ($\textit{max}$: $\NN$) ($\textit{inc}$: stream $\BB$): stream $\NN$ = \\
    \> $@rec@~\rawbind{\textit{count}}{}$ \\
    \> \> @let@ $\textit{count'}$ @=@ $(0~@fby@~\textit{count}) + (@if@~\textit{inc}~@then@~1~@else@~0)$ @in@ \\
    \> \> $\xcheckP{\PSUnknown}{(0 \le \textit{count} \le \textit{max})}$; \\
    \> \> @if @ $\textit{count'} \ge \textit{max}$ @then@ $\textit{max}$  @else@ $\textit{count'}$ \\
\end{tabbing}

The implementation of \emph{count_when} first defines a recursive stream, \emph{count}, which in turn defines a local incremented stream \emph{count'}.
The syntax $0~@fby@~\textit{count}$ can be read as the initial value of zero \emph{followed by} the previous count.

The syntax $\xcheckP{\PSUnknown}{(0 \le \textit{count} \le \textit{max})}$ asserts that the count is within the range $[0, \textit{max}]$.
The subscript $\PSUnknown$ on the check is the \emph{property status}, which in this case denotes that the assertion has been stated, but it is not yet known whether it holds.
A property status of $\PSValid$, on the other hand, denotes that a property has been proved to hold.
These property statuses are used to defer checking properties until enough is known about the environment, and to avoid rechecking properties that have already been proven.
In practice, the user does not explicitly specify property statuses in the source language.
The stated property $(0 \le \textit{count} \le \textit{max})$ is a stream of booleans which must always be true.
Non-streaming operations such as $\le$ are implicitly lifted to streaming operations, and non-streaming values such as $0$ and $\textit{max}$ are impliciitly lifted to constant streams.

We need to defer the proof of the property here, because at the point of stating the property we don't yet have a concrete definition for the count variable.
In this case, we could have instead deferred the \emph{statement} of the property by introducing a let-binding for the count and putting the @check@ outside of the @rec@ combinator.
However, it is not always possible to defer property statements: for example, when calling other streaming functions that have their own preconditions, it may not be possible to move the function call outside of its enclosing @rec@.

@@

To actually prove the property above, we use \fstar{}'s tactics and proof automation to translate the program into a transition system and prove the property inductively on the system.
Finally, we \emph{bless} the expression, which marks the properties as valid.
Blessing is an intensional operation: it traverses the expression and updates the internal metadata, but it does not affect the operational semantics.

@@

The subscript 1 in the translation to transition system and blessing operations refers to the fact that the stream function has one parameter.
The pipit_simplify tactic in the assertion performs normalisation-by-evaluation to simplify away the translation to a first-order transition system; \fstar{}'s proof automation can then solve the inductive check directly.
This proof does require some boilerplate; in the future we hope to minimise this.

Callers of count_when can now use the valid variant without needing to re-prove the count-range property.
In a dedicated model-checker such as Kind2 \cite{champion2016kind2} or Lesar \cite{raymond2008synchronous}, this kind of bookkeeping would all be performed under-the-hood.
By embedding Pipit in a general-purpose theorem prover, we move some of the bookkeeping burden onto the user; however, we also reduce the semantic gap between compilation and proofs, as well as allowing manual proof steps when the automated proofs get stuck.
As an embedded language, we also benefit from writing our programs in a rich metalanguage.

For the trigger fetch logic, we will use count_when as the index into the triggers array.
To retrieve the element from the triggers array, we assume a pure, non-streaming \fstar{} function for array indexing.
We use an unsafe indexing function, which we lift to a streaming primitive and wrap in a contract to make it safe:

\begin{tabbing}
  @MM@\= @MM@ \= \kill
  @let@ array_index ($\textit{array}: \mbox{const_array}~\alpha$) ($\textit{index}$: stream $\NN$): stream $\alpha$ = \\
  \> $@contract@_{\PSUnknown}$ \\
  \> \> \{ $0 \le \textit{index} < \mbox{Array.length}~\textit{array}$ \} \\
  \> \> Array.index_unsafe \textit{array} \textit{index} \\
  \> \> \{ $v.~v~=~\mbox{Array.index_unsafe}~\textit{array}~\textit{index}$ \} \\
\end{tabbing}

Contracts consist of a property status, a rely clause, a body, and a guarantee clause.
The rely clause states that the calling context must ensure that every element of the index stream is a valid index.
In turn, the contract guarantees that every element of the result will be the element of the array at the corresponding index.
The contract status of unknown ($\PSUnknown$) denotes that, although the body satisfies the contract, we have not yet shown that the calling context satisfies the rely clause -- such a proof must wait until the contract is instantiated.
This is a somewhat pathological contract, as the guarantee does not abstract away the details of the implementation at all.

We use the unsafe indexing function, rather than a safe variant with a precondition, as we cannot currently lift primitives with preconditions; we discuss future work to remove this limitation in \autoref{s:core:primitive-tables}.

We can now combine the count_when and index stream functions to implement a (much simplified) time-triggered trigger fetch controller.

\begin{tabbing}
  @MM@\= @MM@ \= @let index@ \= \kill
  @let@ trigger_fetch ($\textit{time}$: stream $\NN$): stream trigger = \\
    \> $@rec @ \textit{trigger}.$ \\
    \> \> $@let @ \textit{inc} = \text{false} @ fby @ (\textit{trigger.time_mark} < \textit{time}) @ in@$\\
    \> \> $@let @ \textit{index} = \mbox{count_when}_{\PSValid}~(\text{Array.size}~\textit{trigger_array} - 1)~\textit{inc} @ in@$ \\
    \> \> $\text{array_index}~\textit{trigger_array}~\textit{index}$\\
\end{tabbing}

The trigger_fetch function takes a stream denoting the current time-since-reset in microseconds and returns a stream of the current active trigger.
Each trigger also has a time mark, which denotes when it should start.
To compute the index, we use the count_when function, so that it increments when the \emph{previous} trigger's time is in the past.
We then use the array_index function to look up the new trigger in the constant array of triggers.

Finally, we need to prove that trigger_fetch's contract instantiation is valid.
This proof script is much the same as the previous one:
\begin{tabbing}
  @MM@\= @MM@ \= \kill
  @let@ $\mbox{trigger_fetch}_{\PSValid}$ ($\textit{max}$: $\NN$): stream $\NN$ $\to$ stream trigger = \\
    \> @let@ $\textit{system} = \mbox{System.translate}_1~\text{trigger_fetch}$ @in@ \\
    \> @assert@ (System.inductive_check $\textit{system}$) @by@ (pipit\_simplify ()); \\
    \> $@bless@_1$ trigger_fetch
\end{tabbing}

Since the implementation of trigger_fetch used $\text{count_valid}_{\PSValid}$ instead of count_valid, the proof did not need to re-prove the property about the range of count.

We have only proved a simple indexing property here but have seen the important pieces of checks, contracts and deferred proofs.
In \autoref{s:evaluation} we will extend this small example to the real system, which uses checked machine-integer arithmetic and proves that scheduling is never late.

% \subsection{Program logics and dependent types}

% Although Pipit is embedded in \fstar{}, which itself supports full dependent types and refinement types, we have opted to for a comparatively weakly-typed language.
% Our stream type \emph{stream $\tau$} denotes a (finite prefix of a) stream of elements of simple type $\tau$, but says nothing about how the stream evolves over time.

% We could consider an alternative approach that was closer to an imperative program logic with preconditions and postconditions.
% For example, we could consider a richer type $\stream \tau~ \{ \rawbind{x}{\phi} \}$ which then the result, of type $\tau$, always satisfies stream postcondition $\phi$.
% We could imagine specifying count_when with such a type:

% \begin{tabbing}
%   @val@ count_when ($\textit{max}$: $\NN$) ($\textit{inc}$: stream $\BB$): stream $\{ \top \} ~\NN~ \{ \rawbind{c}{0 \le c \le max} \}$\\
% \end{tabbing}

% An alternative would be to use refinement types \cite{chen2022synchronous} or some kind of program logic with pre- and post-conditions.

% We believe that such type-based approaches are useful for composing programs together, but 
%  the nested-recursive nature of Lustre programs makes such approaches cumbersome.
% In our system, the recursive combinator has type $@rec@: (\stream \alpha \to \stream \alpha) \to \stream \alpha$.
% Consider instead, if we had a refined stream $\stream \alpha \{ \psi \}$

### 2023-12-20 2024pipit/section/30-Core.tex
ok

\input{figures/core-grammar.tex}
\input{figures/core-typing.tex}
\input{figures/core-bigstep.tex}
\input{figures/core-check.tex}

The grammar of Pipit is defined in \autoref{f:core-grammar}.
The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions except reusing closed functions from the \fstar{} meta-language $(x)$.
Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
% The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
Streams can also be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the values from stream $e'$ for subsequent steps.

Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
% To ensure that streams are productive,
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \cite{caspi1995functional} which is not syntax-directed.
Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.

>>>>

\input{figures/core-grammar.tex}

We now introduce the core Pipit language.
Note that this form differs slightly from the surface syntax presented earlier in \autoref{s:motivation}, which used the syntax of the metalanguage \fstar{}, as well as including proofs in \fstar{} itself.

The grammar of Pipit is defined in \autoref{f:core-grammar}.
The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and primitive applications ($p(\ov{e})$).
Most of the expression forms were introduced informally in \autoref{s:motivation} and correspond to the clock-free expressions of Lustre \cite{caspi1995functional}.

Values, variables and primitives are standard.
The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
% Streams can also be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the values from stream $e'$ for subsequent steps.

Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
% To ensure that streams are productive,
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \cite{caspi1995functional} which is not syntax-directed.
Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax by combining the bindings together into a single tuple.

Checked properties and contracts are annotated with their property status.

We support a fixed set of values and primitives.

Streams $V$ are represented as a non-empty sequence of values; streaming history environments $\Sigma$ are streams of heaps.
Types $\tau$ and type environments $\Gamma$ are standard.

\input{figures/core-typing.tex}

We define the typing judgments for Pipit in \autoref{f:core-typing}.
Most of the typing rules are fairly standard for an unclocked Lustre.
The typing judgment $\typing{\Gamma}{e}{\tau}$ denotes that, in an environment of streams $\Gamma$, expression $e$ denotes a stream of type $\tau$.
This core typing judgment differs from the surface syntax used in \autoref{s:motivation}, which used an explicit stream type.
For the core language, we assume that everything is a stream.

For values, we use an auxiliary judgment form $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$.
Likewise, for primitives we use the auxiliary judgment form $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ to denote that primitive $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
Primitives are pure, non-streaming functions.

Rules \textsc{TValue}, \textsc{TVar}, \textsc{TPrim} and \textsc{TLet} are standard.

Rule \textsc{TFby} states that expression $\xfby{v}{e}$ requires both $v$ and $e$ to have equal types; the result is the same type.

Rule \textsc{TRec} states that a recursive stream $\xrec{x}{e}$ has the recursive stream bound inside $e$.
The recursion must also be guarded, in that any recursive references to $x$ are delayed, but this requirement is performed as a separate syntactic check described in \REF{}.

Rule \textsc{TCheck} states that checking a property $\xcheckP{\PStatus}{e}$ requires a boolean property $e$ and returns unit.

Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\eimpl}{\eguar}$ with a body expression of some type $\tau$.
The overall expression has result type $\tau$.
Both rely and guarantee clauses must be boolean expressions.
Additionally, the guarantee clause can refer to the result value by $x$.

\subsection{Dynamic semantics}
\input{figures/core-bigstep.tex}

The dynamic semantics of Pipit are defined in \autoref{f:core-bigstep}.
We present our semantics in a big-step form.
This differs somewhat from traditional \emph{reactive} semantics of Lustre presented in \cite{caspi1995functional}.
We believe that the big-step semantics is better for performing equational reasoning, as it is substitution-based, while the reactive semantics better emphasises the finite-state execution of the system.
For reasoning about the finite-state execution, we instead use transition systems (\autoref{s:transition}).

The judgment form $\bigstep{\Sigma}{e}{v}$ denotes that expression $e$ evaluates to value $v$ under streaming history $\Sigma$.
The streaming history is a stream of heaps.

\subsection{Checked semantics}
\input{figures/core-check.tex}

We define the checked semantics of Pipit in \autoref{f:core-check}.
\TODO{}

\TODO{Define bless function}
